---
title: "Data products - week 3 assignment"
author: "David Tonarini"
date: "05/02/2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Intro slide
```{r prepare, echo = FALSE}
library(ggplot2, warn.conflicts = FALSE)
library(plotly, warn.conflicts = FALSE)

mtcars$am[which(mtcars$am == 0)] <- 'Manual'
mtcars$am[which(mtcars$am == 1)] <- 'Automatic'
mtcars$am <- as.factor(mtcars$am)

p <- plot_ly(mtcars, x = ~wt, y = ~hp, z = ~mpg, color = ~am, colors = c('#BF382A', '#0C4B8E')) %>%
  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'Weight'),
                     yaxis = list(title = 'Gross horsepower'),
                     zaxis = list(title = 'Miles (US) / gallon')))


```
This sample presentation uses the PlotLy package to display an awesome 3d plot.
It displays the consumptions of a car in relationship to it's weight, gross horsepower, and transmission type (automatic or manual).
The mtcars dataset is used.
Use your keyboard arrows to move to the next slide

## 3D Plot

```{r cars, echo = FALSE}
p
```


## Thank you for watching
3D plots are awesome.
This presentation is completed. 
